Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-04-01 jss/EpiCompare-paper/EpiCompare-paper.Rmd
correction of new order of ternary plot in section 2.

\shannon{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations. The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.  Both epidemics have populations of 1000 people with 10 individuals initially infected.  Additionally note that the two reproduction numbers are the same for each epidemic, $R_0 = 2=0.8/0.4 = 0.64/0.32$.  We plot the epidemics with traditional $state$ vs. $time$ plots. In Fig. \ref{fig:different-scales-standard} we show the time-based paths for the $S$, $I$, and $R$ states for the first 15 days of observed data.  In this time-variant view, we may believe that \ben{epidemic} 1 has a larger $R_0$ than \ben{epidemic} 2 because the peak of infection occurs more quickly than in Epidemic 2.  On the other hand, we may believe \ben{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \sout{model}\ben{epidemic} has not yet peaked at time 15.  In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$.

@@

\shannon{NEW} A time-invariant approach to visualizing epidemics, in comparison, allows us to directly compare $R_0$ from a single plot.  In Fig. \ref{fig:different-scales-tern}
we plot the filamental trajectories of the two epidemics in a time-invariant view. We will explain how and why this works shortly. The important takeaway is that in this time-invariant view, it is apparent that these epidemics are on "the same path." In this case, this indicates that two epidemics have the same value of $R_0$.

@@

\shannon{The reason why we can visually compare $R_0$ in Fig. \ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.} A filamental trajectory can be mathematically viewed as a set of points in space that have an ordering, and that all points on the line between these ordered points are also contained in the geometric object. For a SIR epidemic, we can represent the associated filamental trajectory $\psi$ as

@@

$$
\psi_\xi = \left \{(S(\xi(s)), I(\xi(s)), R(\xi(s))): S, I, R \ge 0, S + I + R = N \right \}_{s\in[0,T]}.
$$
Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then \ben{all points in} $\psi$ will lay in a  \ben{two-dimensional triangular plane in three-dimensional space.  \sout{We can then which can be visualize the \shannon{full filamental trajectory} in a two-dimensional ternary plot}  As a result, we can visualize \shannon{the full filamental trajectory} of an epidemic in a single \ben{ternary plot \sout{2d plot and ultimately $R_0$}}.  \shannon{[Shannon says: Show pic here?]}]\footnote{Which?  \shannon{3d space one? But I'm leaning against it now.   3d never looks good in a paper.}}

@@

The filamental trajectories in Fig. \ref{fig:different-scales-tern} seem to overlap, and we may suspect that something is fundamentally linking these two different epidemics together.  Mathematically, we can show this fundamental link turns is $R_0$.   Let our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.  As with the example, assume both models have the same initial values $(S(0), I(0), R(0))$, and let $R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$. And define $a>0$ to be the relative scalar such that $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$.

>>>>

\ben{[Ben says: this paragraph needs to still be looked at. Also I'm not sure why this particular paragraph was changed so much at all. Currentlly, the way we present comparing these epidemic's $R_0$s isn't well grouped/motivated.]} \shannon{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations. The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.  Both epidemics have populations of 1000 people with 10 individuals initially infected.  Additionally note that the two reproduction numbers are the same for each epidemic, $R_0 = 2=0.8/0.4 = 0.64/0.32$.  We plot the epidemics with traditional $state$ vs. $time$ plots\footnote{This sentecne is out of place / doesn't connect with the other sentences.}. In Fig. \ref{fig:different-scales-standard} we show the time-based paths for the $S$, $I$, and $R$ states for the first 15 days of observed data.  In this time-variant view, we may believe that \ben{epidemic} 1 has a larger $R_0$ than \ben{epidemic} 2 because the peak of infection occurs more quickly than in Epidemic 2.  On the other hand, we may believe \ben{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \ben{epidemic} has not yet peaked at time 15.  In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$\footnote{\ben{This sentnce doesn't connect wit previous examples.}}.

@@

\shannon{NEW} A time-invariant approach to visualizing epidemics, in comparison, allows us to directly compare $R_0$ from a single plot.  \ben{For every time point $t$ we have $S(t),I(t), R(t)$, so we can treat epdiemics as a trajectory in this three-dimensional space, as we visual in the left subplot of Figure \ref{fig:different-sacles-tern}. For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size), we can visual these point ina a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern}'s right subplot. \sout{In Fig. \ref{fig:different-scales-tern}
we plot the filamental trajectories of the two epidemics in a time-invariant view. We will explain how and why this works shortly. The important takeaway is that in this time-invariant view, i}I}t is apparent that these epidemics are on "the same path." In this case, this indicates that two epidemics have the same value of $R_0$.

@@

\ben{Underlying our time-invariant visualization that allowed for the comparison of $R_0$ in Fig. \ref{fig:different-scales-tern} is the treatment of the epidemic as a single filamental trajectory in the state space. \sout{The reason why we can visually compare $R_0$ in Fig. \ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.}} A filamental trajectory can be mathematically viewed as a set of points in space that have an ordering, and that all points on the line between these ordered points are also contained in the geometric object. For a SIR epidemic, we can represent the associated filamental trajectory $\psi$ as

@@

$$
\psi_\xi = \left \{(S(\xi(s)), I(\xi(s)), R(\xi(s))): S, I, R \ge 0, S + I + R = N \right \}_{s\in[0,T]}.
$$
\ben{[Ben says: removed this paragraph now]} Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then \ben{all points in} $\psi$ will lay in a  \ben{two-dimensional triangular plane in three-dimensional space.  \sout{We can then which can be visualize the \shannon{full filamental trajectory} in a two-dimensional ternary plot}  As a result, we can visualize \shannon{the full filamental trajectory} of an epidemic in a single \ben{ternary plot \sout{2d plot and ultimately $R_0$}}.  \shannon{[Shannon says: Show pic here?]}]\footnote{Which?  \shannon{3d space one? But I'm leaning against it now.   3d never looks good in a paper.}}

@@

\ben{As suggested a few paragraphs back, t}he filamental trajectories in Fig. \ref{fig:different-scales-tern} seem to overlap, and we may suspect that something is fundamentally linking these two different epidemics together.  Mathematically, we can show this fundamental link turns is $R_0$.   Let our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.  As with the example, assume both models have the same initial values $(S(0), I(0), R(0))$, and let $R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$. And define $a>0$ to be the relative scalar such that $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$.

### 2021-04-01 jss/EpiCompare-paper/EpiCompare-paper.tex
correction of new order of ternary plot in section 2.

\textcolor{orange}{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.}
For example, consider two epidemics generated from the Kermack and
McKendrick SIR equations. The first epidemic has parameters
\(\beta_1, \gamma_1 = (0.8,0.4)\) and the second has
\(\beta_2, \gamma_2 = (0.64,0.32)\). Both epidemics have populations of
1000 people with 10 individuals initially infected. Additionally note
that the two reproduction numbers are the same for each epidemic,
\(R_0 = 2=0.8/0.4 = 0.64/0.32\). We plot the epidemics with traditional
\(state\) vs.~\(time\) plots. In Fig.
\ref{fig:different-scales-standard} we show the time-based paths for the
\(S\), \(I\), and \(R\) states for the first 15 days of observed data.
In this time-variant view, we may believe that
\textcolor{violet}{epidemic} 1 has a larger \(R_0\) than
\textcolor{violet}{epidemic} 2 because the peak of infection occurs more
quickly than in Epidemic 2. On the other hand, we may believe
\textcolor{violet}{epidemic} 2 has a larger \(R_0\) because its unclear
if the number of infections in that
\sout{model}\textcolor{violet}{epidemic} has not yet peaked at time 15.
In this time-variant view, we cannot determine if one epidemic has
larger value of \(R_0\).

@@

\textcolor{orange}{NEW} A time-invariant approach to visualizing
epidemics, in comparison, allows us to directly compare \(R_0\) from a
single plot. In Fig. \ref{fig:different-scales-tern} we plot the
filamental trajectories of the two epidemics in a time-invariant view.
We will explain how and why this works shortly. The important takeaway
is that in this time-invariant view, it is apparent that these epidemics
are on ``the same path.'' In this case, this indicates that two
epidemics have the same value of \(R_0\).

@@

\textcolor{orange}{The reason why we can visually compare $R_0$ in Fig. \ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.}
A filamental trajectory can be mathematically viewed as a set of points
in space that have an ordering, and that all points on the line between
these ordered points are also contained in the geometric object. For a
SIR epidemic, we can represent the associated filamental trajectory
\(\psi\) as

@@

\[
\psi_\xi = \left \{(S(\xi(s)), I(\xi(s)), R(\xi(s))): S, I, R \ge 0, S + I + R = N \right \}_{s\in[0,T]}.
\] Since the number in each state is non-negative and the sum over the
three states for a given time point sums to \(N\), then
\textcolor{violet}{all points in} \(\psi\) will lay in a
\textbackslash ben\{two-dimensional triangular plane in
three-dimensional space.
\sout{We can then which can be visualize the \textcolor{orange}{full filamental trajectory} in a two-dimensional ternary plot}
As a result, we can visualize
\textcolor{orange}{the full filamental trajectory} of an epidemic in a
single
\textcolor{violet}{ternary plot \sout{2d plot and ultimately $R_0$}}.
\textcolor{orange}{[Shannon says: Show pic here?]}{]}\footnote{Which?  \textcolor{orange}{3d space one? But I'm leaning against it now.   3d never looks good in a paper.}}

@@

The filamental trajectories in Fig. \ref{fig:different-scales-tern} seem
to overlap, and we may suspect that something is fundamentally linking
these two different epidemics together. Mathematically, we can show this
fundamental link turns is \(R_0\). Let our two epidemics be presented as
\(\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}\),
\(\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}\) respectively. As with the
example, assume both models have the same initial values
\((S(0), I(0), R(0))\), and let
\(R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
rate, respectively, for SIR model \(i=1, 2\). And define \(a>0\) to be
the relative scalar such that \(\beta_2 = a \beta_1\) if and only if
\(\gamma_2 = a \gamma_1\).

>>>>

\textcolor{violet}{[Ben says: this paragraph needs to still be looked at. Also I'm not sure why this particular paragraph was changed so much at all. Currentlly, the way we present comparing these epidemic's $R_0$s isn't well grouped/motivated.]}
\textcolor{orange}{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.}
For example, consider two epidemics generated from the Kermack and
McKendrick SIR equations. The first epidemic has parameters
\(\beta_1, \gamma_1 = (0.8,0.4)\) and the second has
\(\beta_2, \gamma_2 = (0.64,0.32)\). Both epidemics have populations of
1000 people with 10 individuals initially infected. Additionally note
that the two reproduction numbers are the same for each epidemic,
\(R_0 = 2=0.8/0.4 = 0.64/0.32\). We plot the epidemics with traditional
\(state\) vs.~\(time\)
plots\footnote{This sentecne is out of place / doesn't connect with the other sentences.}.
In Fig. \ref{fig:different-scales-standard} we show the time-based paths
for the \(S\), \(I\), and \(R\) states for the first 15 days of observed
data. In this time-variant view, we may believe that
\textcolor{violet}{epidemic} 1 has a larger \(R_0\) than
\textcolor{violet}{epidemic} 2 because the peak of infection occurs more
quickly than in Epidemic 2. On the other hand, we may believe
\textcolor{violet}{epidemic} 2 has a larger \(R_0\) because its unclear
if the number of infections in that \textcolor{violet}{epidemic} has not
yet peaked at time 15. In this time-variant view, we cannot determine if
one epidemic has larger value of
\(R_0\)\footnote{\textcolor{violet}{This sentnce doesn't connect wit previous examples.}}.

@@

\textcolor{orange}{NEW} A time-invariant approach to visualizing
epidemics, in comparison, allows us to directly compare \(R_0\) from a
single plot.
\textcolor{violet}{For every time point $t$ we have $S(t),I(t), R(t)$, so we can treat epdiemics as a trajectory in this three-dimensional space, as we visual in the left subplot of Figure \ref{fig:different-sacles-tern}. For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size), we can visual these point ina a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern}'s right subplot. \sout{In Fig. \ref{fig:different-scales-tern}
we plot the filamental trajectories of the two epidemics in a time-invariant view. We will explain how and why this works shortly. The important takeaway is that in this time-invariant view, i}I}t
is apparent that these epidemics are on ``the same path.'' In this case,
this indicates that two epidemics have the same value of \(R_0\).

@@

\textcolor{violet}{Underlying our time-invariant visualization that allowed for the comparison of $R_0$ in Fig. \ref{fig:different-scales-tern} is the treatment of the epidemic as a single filamental trajectory in the state space. \sout{The reason why we can visually compare $R_0$ in Fig. \ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.}}
A filamental trajectory can be mathematically viewed as a set of points
in space that have an ordering, and that all points on the line between
these ordered points are also contained in the geometric object. For a
SIR epidemic, we can represent the associated filamental trajectory
\(\psi\) as

@@

\[
\psi_\xi = \left \{(S(\xi(s)), I(\xi(s)), R(\xi(s))): S, I, R \ge 0, S + I + R = N \right \}_{s\in[0,T]}.
\] \textcolor{violet}{[Ben says: removed this paragraph now]} Since the
number in each state is non-negative and the sum over the three states
for a given time point sums to \(N\), then
\textcolor{violet}{all points in} \(\psi\) will lay in a
\textbackslash ben\{two-dimensional triangular plane in
three-dimensional space.
\sout{We can then which can be visualize the \textcolor{orange}{full filamental trajectory} in a two-dimensional ternary plot}
As a result, we can visualize
\textcolor{orange}{the full filamental trajectory} of an epidemic in a
single
\textcolor{violet}{ternary plot \sout{2d plot and ultimately $R_0$}}.
\textcolor{orange}{[Shannon says: Show pic here?]}{]}\footnote{Which?  \textcolor{orange}{3d space one? But I'm leaning against it now.   3d never looks good in a paper.}}

@@

\textcolor{violet}{As suggested a few paragraphs back, t}he filamental
trajectories in Fig. \ref{fig:different-scales-tern} seem to overlap,
and we may suspect that something is fundamentally linking these two
different epidemics together. Mathematically, we can show this
fundamental link turns is \(R_0\). Let our two epidemics be presented as
\(\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}\),
\(\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}\) respectively. As with the
example, assume both models have the same initial values
\((S(0), I(0), R(0))\), and let
\(R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
rate, respectively, for SIR model \(i=1, 2\). And define \(a>0\) to be
the relative scalar such that \(\beta_2 = a \beta_1\) if and only if
\(\gamma_2 = a \gamma_1\).
